SILVR: Conditional diffusion model for molecule generation without additional training

### 2023-03-15 paper/manuscript.tex
Updates from Overleaf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
- Equivariant diffusion models have been very promising in predicting 3D structures of drug-like molecules. 
- This does not make them suitable candidates for a target binding site. 
- We combine equivariant diffusion models~\cite{huang2022mdm} with Iterative latent variable refinement proposed in Denoising Diffusion probabilistic models~\cite{choi2021ilvr}.
- This allows the generation of new molecules in the shape of the binding site using information from existing fragments.

\section{Theory}
\subsection{Unconditional Equivariant Denoising Diffusion probabilistic models}
\subsubsection*{Denoising Diffusion probabilistic models (DDPM) as generative models}
DDPMs are often used as generative models that were developed for the generation of new images~\cite{}. More recently the same idea has also been applied to molecular generation~\cite{}. The main idea behind diffusion models is for a neural network to learn the reverse of a diffusion process, often referred to as denoising to \textit{sample} a new image or new molecule. In practice this is done by training a neural network $\phi$ and generating samples $\hat{\mathbf{x}}=\psi(X_t,t)$, with $X_t$ with data $X$, noised up to time $t$.  DDPMs consist of two main parts the diffusion process and the denoising process, see the schematic in figure~\ref{fig:diffusion_schematic}.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/model_no_stickman.png}
    \caption{This should show the basic idea of a diffusion model}
    \label{fig:diffusion_schematic}
\end{figure}
The \textit{forward} part of the diffusion model is a Markov process, where you start out with a set of data $\mathbf{x}$ and add noise to this data over a time interval t = [0,\ldots,T] according to the following distribution:
\begin{equation}
  q(\mathbf{x_{1:T}})|\mathbf{x_0} = q(x_0)\prod_{t=1}^T q(\mathbf{x}_t|\mathbf{x}_{t-1})
\end{equation}.
The product of conditional probabilities $q(..)$ can be modelled as a Gaussian Kernel given by:
\begin{equation}
    q(x_t, x_s)=\mathcal{N}(x_t|\alpha_t/\alpha_s, \sigma_t^2-\frac{\alpha_t}{\alpha_s}\sigma^2_s
\end{equation},
for any $t>s$. The parameters $\alpha_t \in \mathbb{R}^{+}$ contains the amount of retained signal and $\sigma_t \in \mathbb{R}^+$ represents the variance and thus the amount of white noise added. 
Different people have looked at different noise schedules~\cite{sohl-dickstein2015deep, ho2020denoising}.
% Next talk about the denoising step
$\hat{p}$ is the parametric probability distribution we want to learn. 
\subsubsection*{Euquivariant diffusion model}
Unlike in image problems, the orientation of molecules matters under translations and rotations. For this purpose, we chose to equivariant diffusion model (EDM) by Hoogeboom et al.~\cite{} as our baseline generative model. The basic concept behind equivarience is that the model is invariant to rotations and translation in this case the E(3) group. This means that scalar (features such as atom types) and vector node properties (such as the positions) will be invariant to group transformations. This means that the order in which a rotation is applied does not matter. You can rotate the input to the model then apply the diffusion/denoising process and get a structure out, or you can apply the diffusion/denoising process first and then do the same rotation to get the same output. 
Mathematically this means that if we have a set of point $\mathbf{x} = (\mathbf{x}_1,\ldots,\mathbf{x}_N) \in \mathbb{^{N\times 3}}$ and each of these points has a set of scalar feature vectors $\mathbf{h}\in \mathbb{R}^{N\times k}$ associated with it these features are invariant to group transformations. The positions translations and rotations are defined according to the orthogonal matrix: $\mathbf{Rx + t} = (\mathbf{R_{x_1}+t},\ldots, \mathbf{R_{x_N}+t}$.
%%%%
% Subsection: ILVR
%%%%
\subsection{Iterative Latent Variable Refinement as a conditioning tool}
Choi et al.~\cite{choi2021ilvr} introduced a way to condition DDPMs by sampling their images from the conditional distribution $p(x_0|c)$, with the condition c:
\begin{equation}
    P_\theta(x_0|c) = \int p_\theta(x_{0:T}|c)Dx_{1:T}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/minimal_model.png}
    \caption{Minimal architecture figure - I prefer this style as it fits better with the theme}
    \label{fig:fig_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/molecules_silvr_rates_horizontal.png}
    \caption{This looks much better than I thought}
    \label{fig:fig_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/silvr_plots.png}
    \caption{Preliminary SILVR results (labels missing). A) Ratio of stable atoms by considering bond distances (note this includes protons but strictly maybe they should be removed. I would have to make new code for that I think). B) RMSD between samples and reference coordinated - pairwise mapping (samples rotated and translated to minimise RMSD). C)Openeye Shapegauss - still concerned I have run this wrong - lower value indicates better shape agreement with receptor D) Geometry stability - samples converted to smiles, geometry predicted using Auto3D, then RMSD with sample}
    \label{fig:fig_3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/fragment_sample_bound.png}
    \caption{Examples of other reference sets and samples fitting binding site}
    \label{fig:fig_5}
\end{figure}

\begin{figure}
    \centering
    \includesvg{Figures/silvr_model_svg.svg}
    \caption{SVG of SILVR model - text not correctly aligned. Opinion vs previous figure.}
    \label{fig:fig_6}
\end{figure}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
- Equivariant diffusion models have been very promising in predicting 3D structures of drug-like molecules. 
- This does not make them suitable candidates for a target binding site. 
- We combine equivariant diffusion models~\cite{huang2022mdm} with Iterative latent variable refinement proposed in Denoising Diffusion probabilistic models~\cite{choi2021ilvr}.
- This allows the generation of new molecules in the shape of the binding site using information from existing fragments.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig1/Schematic.png}
    \caption{Schematic of the equivariant diffusion model with  SILVR indicated for every denoising step.}
    \label{fig:schematic}
\end{figure}
\section{Theory}
\subsection{Unconditional Equivariant Denoising Diffusion probabilistic models}
\subsubsection*{Denoising Diffusion probabilistic models (DDPM) as generative models}
DDPMs are often used as generative models that were developed for the generation of new images~\cite{}. More recently the same idea has also been applied to molecular generation~\cite{}. The main idea behind diffusion models is for a neural network to learn the reverse of a diffusion process, often referred to as denoising to \textit{sample} a new image or new molecule. In practice this is done by training a neural network $\phi$ and generating samples $\hat{\mathbf{x}}=\psi(X_t,t)$, with $X_t$ with data $X$, noised up to time $t$.  DDPMs consist of two main parts the diffusion process and the denoising process, see the schematic in figure~\ref{fig:diffusion_schematic}.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/model_no_stickman.png}
    \caption{This should show the basic idea of a diffusion model}
    \label{fig:diffusion_schematic}
\end{figure}
The \textit{forward} part of the diffusion model is a Markov process, where you start out with a set of data $\mathbf{x}$ and add noise to this data over a time interval t = [0,\ldots,T] according to the following distribution:
\begin{equation}
  q(\mathbf{x_{1:T}})|\mathbf{x_0} = q(x_0)\prod_{t=1}^T q(\mathbf{x}_t|\mathbf{x}_{t-1})
\end{equation}.
The product of conditional probabilities $q(..)$ can be modelled as a Gaussian Kernel given by:
\begin{equation}
    q(x_t, x_s)=\mathcal{N}(x_t|\alpha_t/\alpha_s, \sigma_t^2-\frac{\alpha_t}{\alpha_s}\sigma^2_s
\end{equation},
for any $t>s$. The parameters $\alpha_t \in \mathbb{R}^{+}$ contains the amount of retained signal and $\sigma_t \in \mathbb{R}^+$ represents the variance and thus the amount of white noise added. 
Different people have looked at different noise schedules~\cite{sohl-dickstein2015deep, ho2020denoising}.
% Next talk about the denoising step
$\hat{p}$ is the parametric probability distribution we want to learn. 
\subsubsection*{Euquivariant diffusion model}
Unlike in image problems, the orientation of molecules matters under translations and rotations. For this purpose, we chose to equivariant diffusion model (EDM) by Hoogeboom et al.~\cite{} as our baseline generative model. The basic concept behind equivarience is that the model is invariant to rotations and translation in this case the E(3) group. This means that scalar (features such as atom types) and vector node properties (such as the positions) will be invariant to group transformations. This means that the order in which a rotation is applied does not matter. You can rotate the input to the model then apply the diffusion/denoising process and get a structure out, or you can apply the diffusion/denoising process first and then do the same rotation to get the same output. 
Mathematically this means that if we have a set of point $\mathbf{x} = (\mathbf{x}_1,\ldots,\mathbf{x}_N) \in \mathbb{^{N\times 3}}$ and each of these points has a set of scalar feature vectors $\mathbf{h}\in \mathbb{R}^{N\times k}$ associated with it these features are invariant to group transformations. The positions translations and rotations are defined according to the orthogonal matrix: $\mathbf{Rx + t} = (\mathbf{R_{x_1}+t},\ldots, \mathbf{R_{x_N}+t}$.
%%%%
% Subsection: ILVR
%%%%
\subsection{Iterative Latent Variable Refinement as a conditioning tool}
Choi et al.~\cite{choi2021ilvr} introduced a way to condition DDPMs by sampling their images from the conditional distribution $p(x_0|c)$, with the condition c:
\begin{equation}
    P_\theta(x_0|c) = \int p_\theta(x_{0:T}|c)Dx_{1:T}
\end{equation}

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig3/fig_3_metrics.png}
    \caption{Preliminary SILVR results (labels missing). A) Ratio of stable atoms by considering bond distances (note this includes protons but strictly maybe they should be removed. I would have to make new code for that I think). B) RMSD between samples and reference coordinated - pairwise mapping (samples rotated and translated to minimise RMSD). C)Openeye Shapegauss - still concerned I have run this wrong - lower value indicates better shape agreement with receptor D) Geometry stability - samples converted to smiles, geometry predicted using Auto3D, then RMSD with sample}
    \label{fig:metrics}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Figures/fragment_sample_bound.png}
    \caption{Examples of other reference sets and samples fitting binding site}
    \label{fig:samples}
\end{figure}
